An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-06 EconGrowth.Rnw
Added new "to-do" item about u table and graph.
Added Martin as an author.
Moved tables to an appendix. Added a function that tests whether data is available. Added a parameters table for Cobb-Douglas with U.

\usepackage{soul} %Provides strikethrough text

@@

\author[Calvin]{Caleb Reese}
\author[Calvin]{Lucas Timmer}
\author[Calvin]{Matthew Kuperus Heun\corref{cor1}}
\ead{mkh2@calvin.edu, tel: +1 (616) 526-6663, fax: +1 (616) 526-6501}

\cortext[cor1]{Corresponding author}
\address[Calvin]{Engineering Department, Calvin College, Grand Rapids, MI 49546, USA}

@@

**************** To Do List ********************
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item Heun: Add covariance metrics.
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email but have not heard a response.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item Heun: Add CES production function fits
\item Heun: Add Linex production function fits
\item Heun: Add $k$, $l$, $q$, $x$, and $u$ ONLY fits
\end{itemize}
*************************************************

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(tikzDevice) #See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=FALSE)
opts_chunk$set(
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  warning=TRUE, #Tells whether to show warnings.
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<Load Data>>=
#################################
# This function loads data given a country name.
# The file name from which data will be loaded is assumed to be of the form 
# "<countryName>Data.txt". The file is assumed to exist in a subfolder of this project called "data".
#
# countryName is a string containing the 2-letter abbreviation for the country, e.g. "US" or "CN"
#
# returns a matrix with the data that has been loaded
##
loadData <- function(countryName){
  # Read the data file as a table with a header.  
  fileName <- paste("data/", countryName, "Data.txt", sep="")
  dataTable <- read.table(fileName, header = TRUE)
  # Identifies the header names associated with dataTable
  #print(names(dataTable))
  return(dataTable)
}
@

@@

<<cobb-douglas functions>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
  if (missing(energyType)){
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    modelCD <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha), 
                   data=dataTable, 
                   start = list(lambda=lambdaGuess, alpha=alphaGuess),
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   #upper = list(lambda=Inf, alpha=1)
                   )
    return(modelCD)
  } else {
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    energyColumnName <- paste("i", energyType, sep="") #Energy term desired in the Cobb-Douglas equation.
    eColIndex <- which(names(dataTable) %in% energyColumnName) #Find index of desired column
    colnames(dataTable)[eColIndex] <- "iEToFit" #Change desired column name to "iEToFit"
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    # Now actually do the fit, using the column name "iEToFit".
    modelCDe <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b)), 
                    data = dataTable,
                    start = list(lambda=lambdaGuess, a=aGuess, b=bGuess),
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=FALSE, #Tells whether to print details of curve fit process.
                                          warnOnly=TRUE),
                    #Include the next 3 lines to fit with constraints.
                    #algorithm = "port",
                    #lower = list(lambda=-Inf, a=0, b=0),
                    #upper = list(lambda= Inf, a=1, b=1)
                    )
    return(modelCDe)
  }
}

#########################
# Takes the Cobb-Douglas fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cobbDouglasPredictions <- function(countryAbbrev, energyType){
  # Can't make predictions for any of CN, ZA, SA, IR, TZ, or ZM if we're interested in U
  if (!missing(energyType)){
    # Consider this replacement only if energyType has been specified.
    if (energyType == "U" && (countryAbbrev == "CN" || countryAbbrev == "ZA" || countryAbbrev == "SA" 
                              || countryAbbrev == "IR" || countryAbbrev == "TZ" || countryAbbrev == "ZM")){
      #Return a column of NAs if the above conditions have been met.
      nRows <- 21 # All of these countries need 21 rows.
      df <- as.data.frame(matrix(NA, ncol = 1, nrow = nRows))
      colnames(df) <- "pred"
      return(df)
    }
  }
  model <- cobbDouglasModel(countryAbbrev, energyType)
  pred <- predict(model) #See http://stackoverflow.com/questions/9918807/how-get-plot-from-nls-in-r
  df <- data.frame(pred)
  # Pad with rows as necessary
  df <- padRows(countryAbbrev, df)
  return(df)
}

@@

# returns a vector of data for the Cobb-Douglas model. 
# First item is the +95% CI on all parameters
# Second item contains the parameter estimates
# Third item is the -95% CI on all parameters
# Each row has names: lambda, alpha, beta, gamma, corresponding to the parameters in the model.

@@

cobbDouglasData <- function(countryName, energyType){
  modelCD <- cobbDouglasModel(countryName, energyType)

@@

    lowerCD <- data.frame(lambda=ciCD["lambda","2.5%"], alpha=ciCD["a", "2.5%"], beta=betaCICD[1], gamma=gammaCICD[1])

@@

    upperCD <- data.frame(lambda=ciCD["lambda","97.5%"], alpha=ciCD["a", "97.5%"], beta=betaCICD[2], gamma=gammaCICD[2])

@@

  # Now create the data for a table.

@@

<<CD_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableNoEnergy(), caption.placement="top", sanitize.colnames.function = identity, size="\\tiny")
@

@@

<<CDq_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("Q"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny")
@

@@

<<CDq_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.", eval=TRUE>>=
createCDParamsGraph("Q")
@

@@

<<CDx_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.", eval=TRUE>>=
createCDParamsGraph("X")
@

@@

<<CDx_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("X"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny")
@

@@

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

>>>>

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text

@@

\author[Calvin]{Caleb Reese}
\author[Calvin]{Lucas Timmer}
\author[Stellenbosch]{Martin de Wit}
\author[Calvin]{Matthew Kuperus Heun\corref{cor1}}
\ead{mkh2@calvin.edu, tel: +1 (616) 526-6663, fax: +1 (616) 526-6501}

\cortext[cor1]{Corresponding author}
\address[Calvin]{Engineering Department, Calvin College, Grand Rapids, MI 49546, USA}
\address[Stellenbosch]{ Stellenbosch University, School of Public Leadership, P.O. Box 610, Bellville 7535, South Africa}

@@

**************** To Do List ********************
\begin{itemize}
\item Reese: Add text from Word verson of paper to the LaTeX version.
\item \st{Heun: Create a table of AIC values for each fit.}
\item \st{Heun: Add $u$ predictions for Cobb-Douglas}
\item Heun: Add $u$ parameter table and graph
\item Heun: Add covariance metrics.
\item Heun: Add acknowledgements (Dad, Pruim)
\item Heun: Eliminate blanks in the coefficient tables for the 95\% CIs in the Cobb-Douglas with energy rows. Asked Pruim about this via email but have not heard a response.
\item Heun: move tables to an appendix at the end.
      \begin{itemize}
      \item ZA: lower bound on $\lambda$ and upper bound on $\alpha$
      \item ZM: lower bound on $\lambda$
      \end{itemize}
\item Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.
\item Heun: Add CES production function fits
\item Heun: Add Linex production function fits
\item Heun: Add $k$, $l$, $q$, $x$, and $u$ ONLY fits
\end{itemize}
*************************************************

@@

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(tikzDevice) #See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=FALSE)
opts_chunk$set(
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  eval=TRUE,
  warning=TRUE, #Tells whether to show warnings.
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<Data, eval=TRUE>>=
#################################
# This function loads data given a country abbreviation.
# The file name from which data will be loaded is assumed to be of the form 
# "<countryAbbrev>Data.txt". The file is assumed to exist in a subfolder of this project called "data".
#
# countryAbbrev is a string containing the 2-letter abbreviation for the country, e.g. "US" or "CN"
#
# returns a data.frame with the data that has been loaded
##
loadData <- function(countryAbbrev){
  # Read the data file as a table with a header.  
  fileName <- paste("data/", countryAbbrev, "Data.txt", sep="")
  dataTable <- read.table(fileName, header = TRUE)
  # Identifies the header names associated with dataTable
  #print(names(dataTable))
  return(dataTable)
}

###############################
# Tells whether we have data for this combination of country and energy type.
# a missing energyType or energyType == NA means that we want to develop a model without energy, if possible.
#
# returns logical value (TRUE or FALSE)
##
haveData <- function(countryAbbrev, energyType){
  # The design here is to go through several cases where data is unavailable and
  # return FALSE. If we make it through all of those cases, we'll return TRUE.
  if (!(countryAbbrev %in% countryAbbrevs)){
    # If we don't know the country, we don't have data
    return(FALSE)
  }
  if (missing(energyType) || is.na(energyType)){
    # If we don't want energy included, we have data for all known countries. We need to check this case
    # first, BEFORE doing various checks on the energyType.
    return(TRUE)
  }
  if (!(energyType %in% energyTypes)){
    # If we want to include energy, we know only a few types. So, return false if we don't know the energy type.
    return(FALSE)
  }
  if (energyType == "U" && (!(countryAbbrev %in% countryAbbrevsU))){
    # We have U data for only a few countries.
    return(FALSE)
  }
  return(TRUE)
}
@

@@

<<cobb-douglas functions, eval=TRUE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
  if (missing(energyType)){
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    modelCD <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha), 
                   data=dataTable, 
                   start = list(lambda=lambdaGuess, alpha=alphaGuess),
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   #upper = list(lambda=Inf, alpha=1)
                   )
    return(modelCD)
  } else {
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    energyColumnName <- paste("i", energyType, sep="") #Energy term desired in the Cobb-Douglas equation.
    eColIndex <- which(names(dataTable) %in% energyColumnName) #Find index of desired column
    colnames(dataTable)[eColIndex] <- "iEToFit" #Change desired column name to "iEToFit"
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    # Now actually do the fit, using the column name "iEToFit".
    modelCDe <- nls(iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b)), 
                    data = dataTable,
                    start = list(lambda=lambdaGuess, a=aGuess, b=bGuess),
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=FALSE, #Tells whether to print details of curve fit process.
                                          warnOnly=TRUE),
                    #Include the next 3 lines to fit with constraints.
                    #algorithm = "port",
                    #lower = list(lambda=-Inf, a=0, b=0),
                    #upper = list(lambda= Inf, a=1, b=1)
                    )
    return(modelCDe)
  }
}

#########################
# Takes the Cobb-Douglas fitted models and creates per-country predictions for them.
# Returns a data.frame with the predictions.
##
cobbDouglasPredictions <- function(countryAbbrev, energyType){
  # Can't make predictions for any of CN, ZA, SA, IR, TZ, or ZM if we're interested in U
  if (!missing(energyType)){
    # Consider this replacement only if energyType has been specified.
    if (!(haveData(countryAbbrev, energyType))){
      #If we don't have data for this combination of countryAbbrev and energyType, 
      # return a column of NAs if the above conditions have been met.
      nRows <- 21 # All of these countries need 21 rows.
      df <- as.data.frame(matrix(NA, ncol = 1, nrow = nRows))
      colnames(df) <- "pred"
      return(df)
    }
  }
  model <- cobbDouglasModel(countryAbbrev, energyType)
  pred <- predict(model) #See http://stackoverflow.com/questions/9918807/how-get-plot-from-nls-in-r
  df <- data.frame(pred)
  # Pad with rows as necessary
  df <- padRows(countryAbbrev, df)
  return(df)
}

@@

# returns a data.frame of data for the Cobb-Douglas model. 
# First row is the +95% CI on all parameters
# Second row contains the parameter estimates
# Third row is the -95% CI on all parameters
# Each column has names: lambda, alpha, beta, gamma, corresponding to the parameters in the model.

@@

cobbDouglasData <- function(countryAbbrev, energyType){
  #First, check to see if we want useful work (U) AND one of the countries for which we don't have data.
  if (!haveData(countryAbbrev, energyType)){
    #Return a column of NAs if the above conditions have been met.
    nRows <- 3 # +95% CI, CDe, and -95% CI.
    nCols <- 4 # lambda, alpha, beta, and gamma
    df <- as.data.frame(matrix(NA, ncol = nCols, nrow = nRows))
    colnames(df) <- c("lambda", "alpha", "beta", "gamma")
    rownames(df) <- c("+95% CI", "CDe", "-95% CI")
    return(df)
  # We have a combination of country and energy type for which we have data.
  modelCD <- cobbDouglasModel(countryAbbrev, energyType)

@@

    lowerCD <- data.frame(lambda=ciCD["lambda","2.5%"], 
                          alpha=ciCD["a", "2.5%"], 
                          beta=betaCICD[1], 
                          gamma=gammaCICD[1])

@@

    # The [2] subscripts pick off the upper confidence interval
    upperCD <- data.frame(lambda=ciCD["lambda","97.5%"], 
                          alpha=ciCD["a", "97.5%"], 
                          beta=betaCICD[2], 
                          gamma=gammaCICD[2])

@@

  # Now create the data for a table and return it

@@

@@

@@

<<CDq_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $q$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("Q")
@

@@

<<CDx_Params_Graph, fig.width=figWidth, fig.height=5, fig.cap="Cobb-Douglas (with $x$) model parameters. Vertical bars indicate 95\\% confidence intervals.">>=
createCDParamsGraph("X")
@

@@

@@

\section{Conclusion}

****************** Add conclusion here. *********************

\section{Acknowledgements}

The authors thank Randall Pruim (Calvin College) and Loren L. Heun (Western Michigan University) for their insightful comments on and assistance with the statistical analyses presented herein.

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix
%% \section{}
%% \label{}

\appendix
\section{Tabular Results}

<<CD_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableNoEnergy(), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny", 
      table.placement="H")
@

<<CDq_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("Q"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@

Table \ref{tab:CD_Parameters_With_X} gives the parameters for the Cobb-Douglas model with exergy ($x$).

<<CDx_Params_Table, results='asis'>>=
print(cobbDouglasParamsTableWithEnergy("X"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@

The Cobb-Douglas (with useful work, $u$) parameters are given in Table \ref{tab:CD_Parameters_With_U}.

<<CDu_Params_Table, results='asis', eval=TRUE>>=
print(cobbDouglasParamsTableWithEnergy("U"), 
      caption.placement="top", 
      sanitize.colnames.function = identity, 
      size="\\tiny",
      table.placement="H")
@
